---
title: "simulation conditions model with matrices"
author: "Anne Giacobello"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(rstan)
library(dplyr)
```

```{r simulate true values}
set.seed(42)
# true parameter values
I <- 30; K <- 2; J_per <- 100

true_mu_nu    <- 1.0;  true_b1_nu <- 0.3; true_b2_nu <- 0.5
true_mu_alpha <- 1.5;  true_b_alpha <- 0.2
true_mu_tau   <- 0.3;  true_b_tau   <- 0.02
true_sigma_nu <- 0.3;  true_sigma_alpha <- 0.25; true_sigma_tau <- 0.15

# participant-level covariate
x <- rnorm(I)

# condition dummy: 0 for k=1 und 1 for k=2
cond_dummy <- c(0, 1)

# simulate individual parameters and summary statistics
J_arr <- array(J_per, dim = c(I, K))
C_arr <- array(NA,    dim = c(I, K))
MRT_mat <- array(NA, dim = c(I, K))
VRT_mat <- array(NA, dim = c(I, K))
for (i in 1:I) {
  for (k in 1:K) {
    mean_nu    <- true_mu_nu    + x[i] * true_b1_nu + cond_dummy[k] * true_b2_nu
    mean_alpha <- true_mu_alpha + cond_dummy[k] * true_b_alpha
    mean_tau   <- true_mu_tau   + cond_dummy[k] * true_b_tau
    nu_i    <- rnorm(1, mean_nu,    true_sigma_nu)
    alpha_i <- rnorm(1, mean_alpha, true_sigma_alpha)
    tau_i   <- rnorm(1, mean_tau,   true_sigma_tau)
    
    # bounds
    nu_i    <- max(-3,   min(3,   nu_i))
    alpha_i <- max(0.1,  min(3.0, alpha_i))
    tau_i   <- max(0.05, tau_i)
    
    u      <- exp(-alpha_i * nu_i)
    pi_c   <- 1 / (1 + u)
    mu_rt  <- tau_i + (alpha_i / (2 * nu_i)) * ((1 - u) / (1 + u))
    sig2rt <- (alpha_i / (2 * nu_i^3)) *
              ((1 - 2 * alpha_i * nu_i * u - u^2) / (1 + u)^2)
    
    C_i        <- rbinom(1, J_per, pi_c)
    C_arr[i,k] <- max(C_i, 2)   # need at least 2 for VRT
    MRT_mat[i,k] <- rnorm(1, mu_rt,  sqrt(sig2rt / C_arr[i,k]))
    VRT_mat[i,k] <- rnorm(1, sig2rt, sqrt(2 * sig2rt^2 / (C_arr[i,k] - 1)))
    VRT_mat[i,k] <- max(VRT_mat[i,k], 1e-6)
    J_arr[i,k]   <- J_per
  }
}

# design matrices
X_nu    <- array(NA, dim = c(I, K, 3))
X_alpha <- array(NA, dim = c(I, K, 2))
X_tau   <- array(NA, dim = c(I, K, 2))
for (i in 1:I) for (k in 1:K) {
  X_nu[i,k,]    <- c(1, x[i], cond_dummy[k])
  X_alpha[i,k,] <- c(1, cond_dummy[k])
  X_tau[i,k,]   <- c(1, cond_dummy[k])
}

```


```{r create data}
stan_data <- list(
  I = I, K = K,
  P_nu = 3, P_alpha = 2, P_tau = 2,
  J = J_arr, C = C_arr,
  MRT = MRT_mat, VRT = VRT_mat,
  X_nu = X_nu, X_alpha = X_alpha, X_tau = X_tau
)
```



```{r fit model}
ez_cond <- stan_model("model_w_cond_matrices.stan")
fit <- sampling(ez_cond, data = stan_data,
                chains = 4, iter = 2000, warmup = 1000,
                cores = 4, refresh = 0)
```

```{r recovery check}
# recovery
print(fit, pars = c("beta_nu", "beta_alpha", "beta_tau",
                    "sigma_nu", "sigma_alpha", "sigma_tau"),
      digits = 2)
```

```{r recovery check mit figures}
library(bayesplot)

# extract posterior draws
posterior <- as.array(fit)

# parameters to focus on for convergence checks
pars <- c("beta_nu[1]", "beta_nu[2]", "beta_nu[3]",
          "beta_alpha[1]", "beta_alpha[2]",
          "beta_tau[1]", "beta_tau[2]",
          "sigma_nu", "sigma_alpha", "sigma_tau")

# trace plots
mcmc_trace(posterior, pars = pars)

# R-hat plot, should be close to 1 
rhats <- rhat(fit, pars = pars)
mcmc_rhat(rhats)

# neff should be > 0.1
ratios <- neff_ratio(fit, pars = pars)
mcmc_neff(ratios)

```


```{r bigger simulation}
library(rstan)
library(bayesplot)
library(ggplot2)
library(dplyr)
library(tidyr)

# design
I_vals <- c(20, 60, 150)
J_vals <- c(20, 60, 150)
N_sim  <- 50
K      <- 2

# true parameter values 
true <- list(
  mu_nu    = 1.0, b1_nu = 0.3, b2_nu = 0.5,
  mu_alpha = 1.5, b_alpha = 0.2,
  mu_tau   = 0.3, b_tau  = 0.02,
  sigma_nu = 0.3, sigma_alpha = 0.25, sigma_tau = 0.15
)

# simulation and fit function
run_one <- function(I, J_per, true, model) {
  x          <- rnorm(I)
  cond_dummy <- c(0, 1)
  J_arr      <- array(J_per, dim = c(I, K))
  C_arr      <- array(NA,    dim = c(I, K))
  MRT_mat    <- array(NA,    dim = c(I, K))
  VRT_mat    <- array(NA,    dim = c(I, K))

  true_nu    <- array(NA, dim = c(I, K))
  true_alpha <- array(NA, dim = c(I, K))
  true_tau   <- array(NA, dim = c(I, K))

  for (i in 1:I) for (k in 1:K) {
    nu_i    <- rnorm(1, true$mu_nu    + x[i]*true$b1_nu + cond_dummy[k]*true$b2_nu, true$sigma_nu)
    alpha_i <- rnorm(1, true$mu_alpha + cond_dummy[k]*true$b_alpha,                  true$sigma_alpha)
    tau_i   <- rnorm(1, true$mu_tau   + cond_dummy[k]*true$b_tau,                    true$sigma_tau)

    nu_i    <- max(-3,   min(3,   nu_i))
    alpha_i <- max(0.1,  min(3.0, alpha_i))
    tau_i   <- max(0.05, tau_i)

    true_nu[i,k]    <- nu_i
    true_alpha[i,k] <- alpha_i
    true_tau[i,k]   <- tau_i

    u      <- exp(-alpha_i * nu_i)
    pi_c   <- 1 / (1 + u)
    mu_rt  <- tau_i + (alpha_i / (2*nu_i)) * ((1 - u) / (1 + u))
    sig2rt <- (alpha_i / (2*nu_i^3)) * ((1 - 2*alpha_i*nu_i*u - u^2) / (1+u)^2)

    C_i          <- max(rbinom(1, J_per, pi_c), 2)
    C_arr[i,k]   <- C_i
    MRT_mat[i,k] <- rnorm(1, mu_rt,  sqrt(sig2rt / C_i))
    VRT_mat[i,k] <- max(rnorm(1, sig2rt, sqrt(2*sig2rt^2 / (C_i-1))), 1e-6)
    J_arr[i,k]   <- J_per
  }

  X_nu    <- array(NA, dim = c(I, K, 3))
  X_alpha <- array(NA, dim = c(I, K, 2))
  X_tau   <- array(NA, dim = c(I, K, 2))
  for (i in 1:I) for (k in 1:K) {
    X_nu[i,k,]    <- c(1, x[i], cond_dummy[k])
    X_alpha[i,k,] <- c(1, cond_dummy[k])
    X_tau[i,k,]   <- c(1, cond_dummy[k])
  }

  fit <- tryCatch(sampling(model,
    data = list(I=I, K=K, P_nu=3, P_alpha=2, P_tau=2,
                J=J_arr, C=C_arr, MRT=MRT_mat, VRT=VRT_mat,
                X_nu=X_nu, X_alpha=X_alpha, X_tau=X_tau),
    chains=2, iter=1500, warmup=500, cores=2, refresh=0),
    error = function(e) NULL)

  if (is.null(fit)) return(NULL)

  summ <- summary(fit)$summary

  # group-level parameters 
  group_results <- data.frame(
    param     = c("mu_nu", "mu_alpha", "mu_tau"),
    true_val  = c(true$mu_nu, true$mu_alpha, true$mu_tau),
    estimate  = summ[c("beta_nu[1]", "beta_alpha[1]", "beta_tau[1]"), "mean"],
    q5        = summ[c("beta_nu[1]", "beta_alpha[1]", "beta_tau[1]"), "2.5%"],
    q95       = summ[c("beta_nu[1]", "beta_alpha[1]", "beta_tau[1]"), "97.5%"],
    condition = NA,
    level     = "group",
    I = I, J = J_per,
    row.names = NULL
  )

  # individual-level parameters per condition
  indiv_results <- bind_rows(lapply(1:K, function(k) {
    nu_names    <- paste0("nu[",    1:I, ",", k, "]")
    alpha_names <- paste0("alpha[", 1:I, ",", k, "]")
    tau_names   <- paste0("tau[",   1:I, ",", k, "]")

    bind_rows(
      data.frame(param     = "nu",
                 true_val  = true_nu[, k],
                 estimate  = summ[nu_names,    "mean"],
                 q5        = summ[nu_names,    "2.5%"],
                 q95       = summ[nu_names,    "97.5%"],
                 condition = k, level = "individual", I = I, J = J_per),
      data.frame(param     = "alpha",
                 true_val  = true_alpha[, k],
                 estimate  = summ[alpha_names, "mean"],
                 q5        = summ[alpha_names, "2.5%"],
                 q95       = summ[alpha_names, "97.5%"],
                 condition = k, level = "individual", I = I, J = J_per),
      data.frame(param     = "tau",
                 true_val  = true_tau[, k],
                 estimate  = summ[tau_names,   "mean"],
                 q5        = summ[tau_names,   "2.5%"],
                 q95       = summ[tau_names,   "97.5%"],
                 condition = k, level = "individual", I = I, J = J_per)
    )
  }))

  bind_rows(group_results, indiv_results)
}

# run all conditions 
set.seed(42)
ez_cond <- stan_model("model_w_cond_matrices.stan")

results <- bind_rows(lapply(I_vals, function(I) {
  bind_rows(lapply(J_vals, function(J) {
    bind_rows(lapply(seq_len(N_sim), function(s) run_one(I, J, true, ez_cond)))
  }))
}))

# prepare for plotting 
results <- results |>
  mutate(
    I_label = factor(paste0("I = ", I), levels = paste0("I = ", I_vals)),
    J_label = factor(paste0("J = ", J), levels = paste0("J = ", J_vals)),
    condition = ifelse(is.na(condition), NA, paste0("Condition ", condition))
  )

results_group <- results |> filter(level == "group")
results_indiv <- results |> filter(level == "individual")

# plot theme
dark_theme <- theme(
  plot.background  = element_rect(fill = "black", colour = NA),
  plot.title       = element_text(colour = "white", size = 13, hjust = 0.5),
  panel.background = element_rect(fill = "black", colour = NA),
  panel.grid.major = element_line(colour = "grey25", linewidth = 0.3),
  panel.grid.minor = element_blank(),
  panel.border     = element_rect(colour = "white", fill = NA, linewidth = 0.4),
  strip.background = element_rect(fill = "black", colour = NA),
  strip.text       = element_text(colour = "white", size = 10),
  axis.text        = element_text(colour = "white", size = 8),
  axis.ticks       = element_line(colour = "white"),
  axis.title       = element_blank(),
  legend.position  = "none"
)

# plot function 
make_plot <- function(df, param_name, label, by_condition = FALSE) {
  p <- ggplot(df |> filter(param == param_name),
              aes(x = true_val, y = estimate)) +
    geom_abline(intercept = 0, slope = 1,
                linetype = "dashed", colour = "white", linewidth = 0.5) +
    geom_smooth(method = "lm", colour = "goldenrod3",
                fill = "goldenrod3", alpha = 0.3, linewidth = 0.7) +
    geom_point(colour = "goldenrod", alpha = 0.6, size = 1.2) +
    labs(title = label) +
    dark_theme

  if (by_condition) {
    p + facet_grid(I_label + condition ~ J_label)
  } else {
    p + facet_grid(I_label ~ J_label)
  }
}

# group-level plots 
group_labels <- list(
  mu_nu    = expression(mu^{(nu)}),
  mu_alpha = expression(mu^{(alpha)}),
  mu_tau   = expression(mu^{(tau)})
)

plots_group <- lapply(names(group_labels), function(p)
  make_plot(results_group, p, group_labels[[p]], by_condition = FALSE))
names(plots_group) <- names(group_labels)

# individual-level plots (split by condition) 
indiv_labels <- list(
  nu    = expression(nu),
  alpha = expression(alpha),
  tau   = expression(tau)
)

plots_indiv <- lapply(names(indiv_labels), function(p)
  make_plot(results_indiv, p, indiv_labels[[p]], by_condition = TRUE))
names(plots_indiv) <- names(indiv_labels)

# view 
plots_group[["mu_nu"]]
plots_group[["mu_alpha"]]
plots_group[["mu_tau"]]

plots_indiv[["nu"]]
plots_indiv[["alpha"]]
plots_indiv[["tau"]]

# save 
for (p in names(plots_group)) {
  ggsave(paste0("recovery_group_", p, ".png"),
         plot = plots_group[[p]], width = 8, height = 7,
         bg = "black", dpi = 300)
}

for (p in names(plots_indiv)) {
  ggsave(paste0("recovery_indiv_", p, ".png"),
         plot = plots_indiv[[p]], width = 8, height = 10,
         bg = "black", dpi = 300)
}

```

```{r testing biases for tau}
# compute bias for individual tau estimates
results_indiv <- results_indiv |>
  mutate(bias = estimate - true_val)

# does bias correlate with the true tau value
ggplot(results_indiv |> filter(param == "tau"),
       aes(x = true_val, y = bias)) +
  geom_point(colour = "goldenrod", alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", colour = "white", fill = "grey40") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "white") +
  facet_grid(I_label + condition ~ J_label) +
  labs(title = "Tau bias vs. true tau (boundary effect test)") +
  dark_theme
# negative slope = boundary pushes estimates at low true values

# does tau bias correlate with nu bias
bias_wide <- results_indiv |>
  filter(param %in% c("tau", "nu")) |>
  select(param, bias, condition, I, J, I_label, J_label) |>
  group_by(param, condition, I, J, I_label, J_label) |>
  mutate(id = row_number()) |>
  pivot_wider(names_from = param, values_from = bias)

ggplot(bias_wide, aes(x = nu, y = tau)) +
  geom_point(colour = "goldenrod", alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", colour = "white", fill = "grey40") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "white") +
  facet_grid(I_label + condition ~ J_label) +
  labs(title = "Tau bias vs. nu bias (confounding test)",
       x = "Nu bias", y = "Tau bias") +
  dark_theme
# negative correlation = tau and nu are compensating each other

# does group-level mu_tau drive individual bias
results_group |>
  filter(param == "mu_tau") |>
  mutate(group_bias = estimate - true_val) |>
  ggplot(aes(x = true_val, y = estimate)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "white") +
  geom_point(colour = "goldenrod", alpha = 0.6) +
  geom_smooth(method = "lm", colour = "goldenrod3", fill = "goldenrod3", alpha = 0.3) +
  facet_grid(I_label ~ J_label) +
  labs(title = expression(paste("Group-level ", mu^{(tau)}, " recovery"))) +
  dark_theme
# if mu_tau is often above the line --> shrinkage

# does bias worsen at high accuracy
results_indiv |>
  filter(param == "tau") |>
  ggplot(aes(x = true_val, y = bias, colour = condition)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "white") +
  scale_colour_manual(values = c("goldenrod", "steelblue")) +
  facet_grid(I_label ~ J_label) +
  labs(title = "Tau bias by condition (accuracy test)") +
  dark_theme
# if condition 1 (lower nu, lower accuracy) shows less bias than condition 2, the high accuracy destabilizes tau
```

```{r new big simulation with tau convergence errors fixed in the sampler}
library(rstan)
library(bayesplot)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)

# design 
I_vals <- c(20, 60, 150)
J_vals <- c(20, 60, 150)
N_sim  <- 50
K      <- 2

# true parameter values
true <- list(
  mu_nu    = 1.0, b1_nu = 0.3, b2_nu = 0.5,
  mu_alpha = 1.5, b_alpha = 0.2,
  mu_tau   = 0.3, b_tau  = 0.02,
  sigma_nu = 0.3, sigma_alpha = 0.25, sigma_tau = 0.15
)

# simulation and fit function 
run_one <- function(I, J_per, true, model,
                    rhat_threshold = 1.05,
                    hopeless_threshold = 10,
                    iter1 = 1500, warmup1 = 500,
                    iter2 = 3000, warmup2 = 1000) {

  x          <- rnorm(I)
  cond_dummy <- c(0, 1)
  J_arr      <- array(J_per, dim = c(I, K))
  C_arr      <- array(NA,    dim = c(I, K))
  MRT_mat    <- array(NA,    dim = c(I, K))
  VRT_mat    <- array(NA,    dim = c(I, K))
  true_nu    <- array(NA,    dim = c(I, K))
  true_alpha <- array(NA,    dim = c(I, K))
  true_tau   <- array(NA,    dim = c(I, K))

  for (i in 1:I) for (k in 1:K) {
    nu_i    <- rnorm(1, true$mu_nu    + x[i]*true$b1_nu + cond_dummy[k]*true$b2_nu, true$sigma_nu)
    alpha_i <- rnorm(1, true$mu_alpha + cond_dummy[k]*true$b_alpha,                  true$sigma_alpha)
    tau_i   <- rnorm(1, true$mu_tau   + cond_dummy[k]*true$b_tau,                    true$sigma_tau)

    nu_i    <- max(-3,   min(3,   nu_i))
    alpha_i <- max(0.1,  min(3.0, alpha_i))
    tau_i   <- max(0.05, tau_i)

    true_nu[i,k]    <- nu_i
    true_alpha[i,k] <- alpha_i
    true_tau[i,k]   <- tau_i

    u      <- exp(-alpha_i * nu_i)
    pi_c   <- 1 / (1 + u)
    mu_rt  <- tau_i + (alpha_i / (2*nu_i)) * ((1 - u) / (1 + u))
    sig2rt <- (alpha_i / (2*nu_i^3)) * ((1 - 2*alpha_i*nu_i*u - u^2) / (1+u)^2)

    C_i          <- max(rbinom(1, J_per, pi_c), 2)
    C_arr[i,k]   <- C_i
    MRT_mat[i,k] <- rnorm(1, mu_rt,  sqrt(sig2rt / C_i))
    VRT_mat[i,k] <- max(rnorm(1, sig2rt, sqrt(2*sig2rt^2 / (C_i-1))), 1e-6)
    J_arr[i,k]   <- J_per
  }

  X_nu    <- array(NA, dim = c(I, K, 3))
  X_alpha <- array(NA, dim = c(I, K, 2))
  X_tau   <- array(NA, dim = c(I, K, 2))
  for (i in 1:I) for (k in 1:K) {
    X_nu[i,k,]    <- c(1, x[i], cond_dummy[k])
    X_alpha[i,k,] <- c(1, cond_dummy[k])
    X_tau[i,k,]   <- c(1, cond_dummy[k])
  }

  stan_data <- list(I=I, K=K, P_nu=3, P_alpha=2, P_tau=2,
                    J=J_arr, C=C_arr, MRT=MRT_mat, VRT=VRT_mat,
                    X_nu=X_nu, X_alpha=X_alpha, X_tau=X_tau)

  # return clean exclusion row
  exclude_row <- function(reason) {
    data.frame(param=NA, true_val=NA, estimate=NA, q5=NA, q95=NA,
               condition=NA, level=NA, excluded=TRUE,
               reason=reason, I=I, J=J_per)
  }

  # true when fit is so degenerate that refitting is useless
  # triggers if any R-hat > hopeless_threshold or if R-hats contain NaN or Inf
  hopeless <- function(fit) {
    rhats <- tryCatch(summary(fit)$summary[, "Rhat"], error = function(e) NA)
    any(is.nan(rhats) | is.infinite(rhats) | rhats > hopeless_threshold,
        na.rm = FALSE)
  }

  # helper: TRUE when all R-hats are below the convergence threshold
  converged <- function(fit) {
    rhats <- summary(fit)$summary[, "Rhat"]
    !any(rhats > rhat_threshold, na.rm = TRUE)
  }

  # first attempt
  fit <- tryCatch(
    sampling(model, data = stan_data,
             chains=2, iter=iter1, warmup=warmup1,
             cores=2, refresh=0,
             control=list(adapt_delta=0.90)),
    error = function(e) NULL)

  if (is.null(fit)) {
    message("First fit crashed — skipping (I=", I, ", J=", J_per, ")")
    return(exclude_row("crash_1"))
  }

  # check for hopeless degeneracy before attempting refit
  if (hopeless(fit)) {
    max_rhat <- round(max(summary(fit)$summary[, "Rhat"], na.rm = TRUE), 3)
    message("Degenerate fit — excluding without refit (max Rhat=", max_rhat,
            ", I=", I, ", J=", J_per, ")")
    return(exclude_row("degenerate"))
  }

  # refit only if rhat_threshold < Rhat <= hopeless_threshold
  if (!converged(fit)) {
    max_rhat <- round(max(summary(fit)$summary[, "Rhat"], na.rm=TRUE), 3)
    message("Refit triggered (max Rhat=", max_rhat,
            ", I=", I, ", J=", J_per, ")")

    fit <- tryCatch(
      sampling(model, data=stan_data,
               chains=4, iter=iter2, warmup=warmup2,
               cores=4, refresh=0,
               control=list(adapt_delta=0.95)),
      error = function(e) NULL)

    if (is.null(fit)) {
      message("Second fit crashed — excluding (I=", I, ", J=", J_per, ")")
      return(exclude_row("crash_2"))
    }

    if (!converged(fit)) {
      max_rhat <- round(max(summary(fit)$summary[, "Rhat"], na.rm=TRUE), 3)
      message("Still not converged after refit (max Rhat=", max_rhat,
              ") — excluding (I=", I, ", J=", J_per, ")")
      return(exclude_row("no_convergence"))
    }
  }

  # extract results 
  summ <- summary(fit)$summary

  group_results <- data.frame(
    param     = c("mu_nu", "mu_alpha", "mu_tau"),
    true_val  = c(true$mu_nu, true$mu_alpha, true$mu_tau),
    estimate  = summ[c("beta_nu[1]", "beta_alpha[1]", "beta_tau[1]"), "mean"],
    q5        = summ[c("beta_nu[1]", "beta_alpha[1]", "beta_tau[1]"), "2.5%"],
    q95       = summ[c("beta_nu[1]", "beta_alpha[1]", "beta_tau[1]"), "97.5%"],
    condition = NA, level = "group", excluded = FALSE, reason = NA,
    I = I, J = J_per, row.names = NULL
  )

  indiv_results <- bind_rows(lapply(1:K, function(k) {
    nu_names    <- paste0("nu[",    1:I, ",", k, "]")
    alpha_names <- paste0("alpha[", 1:I, ",", k, "]")
    tau_names   <- paste0("tau[",   1:I, ",", k, "]")
    bind_rows(
      data.frame(param="nu",    true_val=true_nu[,k],
                 estimate=summ[nu_names,    "mean"],
                 q5=summ[nu_names,    "2.5%"], q95=summ[nu_names,    "97.5%"],
                 condition=k, level="individual", excluded=FALSE, reason=NA,
                 I=I, J=J_per),
      data.frame(param="alpha", true_val=true_alpha[,k],
                 estimate=summ[alpha_names, "mean"],
                 q5=summ[alpha_names, "2.5%"], q95=summ[alpha_names, "97.5%"],
                 condition=k, level="individual", excluded=FALSE, reason=NA,
                 I=I, J=J_per),
      data.frame(param="tau",   true_val=true_tau[,k],
                 estimate=summ[tau_names,   "mean"],
                 q5=summ[tau_names,   "2.5%"], q95=summ[tau_names,   "97.5%"],
                 condition=k, level="individual", excluded=FALSE, reason=NA,
                 I=I, J=J_per)
    )
  }))

  bind_rows(group_results, indiv_results)
}

# run all conditions 
set.seed(42)
ez_cond <- stan_model("model_w_cond_matrices.stan")

all_results <- bind_rows(lapply(I_vals, function(I) {
  bind_rows(lapply(J_vals, function(J) {
    bind_rows(lapply(seq_len(N_sim), function(s) run_one(I, J, true, ez_cond)))
  }))
}))

# split good fits from excluded
excluded <- all_results |> filter(excluded == TRUE)
results  <- all_results |> filter(excluded == FALSE)

# exclusion rate table 
exclusion_table <- excluded |>
  group_by(I, J) |>
  summarise(n_excluded = n(), .groups = "drop") |>
  mutate(pct = round(n_excluded / N_sim * 100, 1)) |>
  pivot_wider(names_from=J, values_from=c(n_excluded, pct),
              names_glue="{.value}_J{J}", values_fill=0)

# cleaner version 
exclusion_report <- excluded |>
  group_by(I, J, reason) |>
  summarise(n = n(), pct = round(n() / N_sim * 100, 1), .groups="drop")

print(exclusion_report)

# numbers for manuscript 
total_sims  <- N_sim * length(I_vals) * length(J_vals)
n_excluded  <- nrow(excluded)
pct_overall <- round(n_excluded / total_sims * 100, 1)

# prepare plotting 
results <- results |>
  mutate(
    I_label   = factor(paste0("I = ", I), levels=paste0("I = ", I_vals)),
    J_label   = factor(paste0("J = ", J), levels=paste0("J = ", J_vals)),
    condition = ifelse(is.na(condition), NA, paste0("Condition ", condition))
  )

results_group <- results |> filter(level == "group")
results_indiv <- results |> filter(level == "individual")

# plot theme
dark_theme <- theme(
  plot.background  = element_rect(fill="black", colour=NA),
  plot.title       = element_text(colour="white", size=13, hjust=0.5),
  panel.background = element_rect(fill="black", colour=NA),
  panel.grid.major = element_line(colour="grey25", linewidth=0.3),
  panel.grid.minor = element_blank(),
  panel.border     = element_rect(colour="white", fill=NA, linewidth=0.4),
  strip.background = element_rect(fill="black", colour=NA),
  strip.text       = element_text(colour="white", size=10),
  axis.text        = element_text(colour="white", size=8),
  axis.ticks       = element_line(colour="white"),
  axis.title       = element_blank(),
  legend.position  = "none"
)

# plot function 
make_plot <- function(df, param_name, label, by_condition=FALSE) {
  p <- ggplot(df |> filter(param == param_name),
              aes(x=true_val, y=estimate)) +
    geom_abline(intercept=0, slope=1,
                linetype="dashed", colour="white", linewidth=0.5) +
    geom_smooth(method="lm", colour="goldenrod3",
                fill="goldenrod3", alpha=0.3, linewidth=0.7) +
    geom_point(colour="goldenrod", alpha=0.6, size=1.2) +
    labs(title=label) +
    dark_theme

  if (by_condition) {
    p + facet_grid(I_label + condition ~ J_label)
  } else {
    p + facet_grid(I_label ~ J_label)
  }
}

# group-level plots 
group_labels <- list(
  mu_nu    = expression(mu^{(nu)}),
  mu_alpha = expression(mu^{(alpha)}),
  mu_tau   = expression(mu^{(tau)})
)

plots_group <- lapply(names(group_labels), function(p)
  make_plot(results_group, p, group_labels[[p]], by_condition=FALSE))
names(plots_group) <- names(group_labels)

# individual-level plots split by condition
indiv_labels <- list(
  nu    = expression(nu),
  alpha = expression(alpha),
  tau   = expression(tau)
)

plots_indiv <- lapply(names(indiv_labels), function(p)
  make_plot(results_indiv, p, indiv_labels[[p]], by_condition=TRUE))
names(plots_indiv) <- names(indiv_labels)

# view 
plots_group[["mu_nu"]]
plots_group[["mu_alpha"]]
plots_group[["mu_tau"]]

plots_indiv[["nu"]]
plots_indiv[["alpha"]]
plots_indiv[["tau"]]

# save 
for (p in names(plots_group)) {
  ggsave(paste0("recovery_group_", p, ".png"),
         plot=plots_group[[p]], width=8, height=7,
         bg="black", dpi=300)
}

for (p in names(plots_indiv)) {
  ggsave(paste0("recovery_indiv_", p, ".png"),
         plot=plots_indiv[[p]], width=8, height=10,
         bg="black", dpi=300)
}
```


```{r report exclusions}
library(knitr)

# reshape
exclusion_table <- exclusion_report |>
  group_by(I, J) |>
  summarise(pct_excluded = sum(pct), .groups = "drop") |>
  pivot_wider(names_from   = J,
              values_from  = pct_excluded,
              names_prefix = "J = ",
              values_fill  = 0) |>
  rename("I" = I)

kable(exclusion_table,
      caption = "Percentage of simulations excluded due to non-convergence 
                 (R-hat > 1.05 after refit) per combination of number of 
                 participants (I) and trials per participant (J).",
      col.names = c("$I$", "J = 20", "J = 60", "J = 150"),
      align = "c")
```

```{r group level estimates}
# group-level bias and coverage 

results_group <- results_group |>
  mutate(
    bias     = estimate - true_val,
    covered  = q5 <= true_val & true_val <= q95
  )

# bias distributions (violin)
plot_group_bias <- function(param_name, label) {
  ggplot(results_group |> filter(param == param_name),
         aes(x = J_label, y = bias)) +
    geom_hline(yintercept = 0, colour = "white",
               linetype = "dashed", linewidth = 0.5) +
    geom_violin(fill = "goldenrod", colour = NA, alpha = 0.35,
                trim = TRUE, scale = "width") +
    geom_jitter(colour = "goldenrod", alpha = 0.7, size = 1.4,
                width = 0.08) +
    facet_grid(I_label ~ .) +
    labs(title = label,
         x = NULL, y = "Bias (estimate \u2212 true)") +
    dark_theme +
    theme(axis.title.y = element_text(colour = "white", size = 9))
}

bias_labels_group <- list(
  mu_nu    = expression(mu^{(nu)}),
  mu_alpha = expression(mu^{(alpha)}),
  mu_tau   = expression(mu^{(tau)})
)

plots_bias_group <- lapply(names(bias_labels_group), function(p)
  plot_group_bias(p, bias_labels_group[[p]]))
names(plots_bias_group) <- names(bias_labels_group)

plots_bias_group[["mu_nu"]]
plots_bias_group[["mu_alpha"]]
plots_bias_group[["mu_tau"]]


# coverage rates 
coverage_group <- results_group |>
  group_by(param, I_label, J_label) |>
  summarise(coverage = mean(covered, na.rm = TRUE) * 100,
            .groups = "drop")

plot_group_coverage <- function(param_name, label) {
  ggplot(coverage_group |> filter(param == param_name),
         aes(x = J_label, y = coverage)) +
    geom_hline(yintercept = 95, colour = "white",
               linetype = "dashed", linewidth = 0.5) +
    geom_col(fill = "goldenrod", alpha = 0.8, width = 0.6) +
    geom_text(aes(label = paste0(round(coverage, 1), "%")),
              colour = "white", vjust = -0.4, size = 3) +
    facet_grid(I_label ~ .) +
    coord_cartesian(ylim = c(0, 105)) +
    labs(title = label,
         x = NULL, y = "95% CI coverage (%)") +
    dark_theme +
    theme(axis.title.y = element_text(colour = "white", size = 9))
}

plots_cov_group <- lapply(names(bias_labels_group), function(p)
  plot_group_coverage(p, bias_labels_group[[p]]))
names(plots_cov_group) <- names(bias_labels_group)

plots_cov_group[["mu_nu"]]
plots_cov_group[["mu_alpha"]]
plots_cov_group[["mu_tau"]]


# save 
for (p in names(plots_bias_group)) {
  ggsave(paste0("bias_group_", p, ".png"),
         plot = plots_bias_group[[p]], width = 5, height = 7,
         bg = "black", dpi = 300)
  ggsave(paste0("coverage_group_", p, ".png"),
         plot = plots_cov_group[[p]], width = 5, height = 7,
         bg = "black", dpi = 300)
}
```

```{r save results}
saveRDS(all_results, file = "simulation_results.rds")
```

```{r reload}
# reload results 
all_results <- readRDS("simulation_results.rds")

# re-run the post-processing block 
excluded <- all_results |> filter(excluded == TRUE)
results  <- all_results |> filter(excluded == FALSE)

exclusion_report <- excluded |>
  group_by(I, J, reason) |>
  summarise(n = n(), pct = round(n() / N_sim * 100, 1), .groups = "drop")

total_sims  <- N_sim * length(I_vals) * length(J_vals)
n_excluded  <- nrow(excluded)
pct_overall <- round(n_excluded / total_sims * 100, 1)

results <- results |>
  mutate(
    I_label   = factor(paste0("I = ", I), levels = paste0("I = ", I_vals)),
    J_label   = factor(paste0("J = ", J), levels = paste0("J = ", J_vals)),
    condition = ifelse(is.na(condition), NA, paste0("Condition ", condition))
  )

results_group <- results |> filter(level == "group")
results_indiv <- results |> filter(level == "individual")
```



-----

matrix[2, P_nu] prior_beta_nu	--> prior_beta_nu = priors$beta_nu (2×3 matrix from derive_priors())
matrix[2, P_alpha] prior_beta_alpha	--> prior_beta_alpha = priors$beta_alpha (2×2 matrix)
matrix[2, P_tau] prior_beta_tau	--> prior_beta_tau = priors$beta_tau (2×2 matrix)
vector[2] prior_sigma_nu -->	prior_sigma_nu = priors$sigma_nu (length-2 vector)
vector[2] prior_sigma_alpha	--> prior_sigma_alpha = priors$sigma_alpha
vector[2] prior_sigma_tau -->	prior_sigma_tau = priors$sigma_tau


```{r new big simulation with priors from data}
library(rstan)
library(bayesplot)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(parallel)

# design 
I_vals <- c(20, 60, 150)
J_vals <- c(20, 60, 150)
N_sim  <- 5 #50
K      <- 2

# true parameter values
true <- list(
  mu_nu    = 1.0, b1_nu = 0.3, b2_nu = 0.5,
  mu_alpha = 1.5, b_alpha = 0.2,
  mu_tau   = 0.3, b_tau  = 0.02,
  sigma_nu = 0.3, sigma_alpha = 0.25, sigma_tau = 0.15
)

# point estimates from observed statistics per participant
ez_point_estimates <- function(C_arr, J_arr, MRT_mat, VRT_mat, I, K) {
  nu_hat    <- matrix(NA, I, K)
  alpha_hat <- matrix(NA, I, K)
  tau_hat   <- matrix(NA, I, K)

  for (i in 1:I) for (k in 1:K) {
    J_ik  <- J_arr[i, k]
    Pc    <- C_arr[i, k] / J_ik
    Pc    <- max(min(Pc, 1 - 1 / (2 * J_ik)), 1 / (2 * J_ik))
    VRT   <- VRT_mat[i, k]
    MRT   <- MRT_mat[i, k]

    L      <- log(Pc / (1 - Pc))
    inside <- L * (Pc^2 * L - Pc * L + Pc - 0.5) / VRT

    if (is.na(inside) || inside <= 0) next

    nu_ik    <- sign(Pc - 0.5) * inside^(1/4)
    if (abs(nu_ik) < 1e-6) next

    alpha_ik <- L / nu_ik
    u        <- exp(-alpha_ik * nu_ik)
    tau_ik   <- MRT - (alpha_ik / (2 * nu_ik)) * ((1 - u) / (1 + u))

    nu_hat[i, k]    <- nu_ik
    alpha_hat[i, k] <- alpha_ik
    tau_hat[i, k]   <- tau_ik
  }

  list(nu = nu_hat, alpha = alpha_hat, tau = tau_hat)
}

# take prior matrices from EZ point estimates
# row 1 = prior means, row 2 = prior SDs
# intercept mean = general mean of EZ estimates
# slope mean = 0 (no direction)
# SDs = 2 x SD EZ estimates 
derive_priors <- function(ez) {
  nu_vals    <- as.vector(ez$nu)
  alpha_vals <- as.vector(ez$alpha)
  tau_vals   <- as.vector(ez$tau)

  nu_vals    <- nu_vals[!is.na(nu_vals) & nu_vals > -3   & nu_vals < 3]
  alpha_vals <- alpha_vals[!is.na(alpha_vals) & alpha_vals > 0.1 & alpha_vals < 3]
  tau_vals   <- tau_vals[!is.na(tau_vals) & tau_vals > 0.05]

  nu_sd    <- sd(nu_vals)
  alpha_sd <- sd(alpha_vals)
  tau_sd   <- sd(tau_vals)

  list(
    beta_nu = matrix(
      c(mean(nu_vals), 0,         0,
        nu_sd * 2,     nu_sd * 2, nu_sd * 2),
      nrow = 2, byrow = TRUE
    ),
    beta_alpha = matrix(
      c(mean(alpha_vals), 0,
        alpha_sd * 2,     alpha_sd * 2),
      nrow = 2, byrow = TRUE
    ),
    beta_tau = matrix(
      c(mean(tau_vals), 0,
        tau_sd * 2,     tau_sd * 2),
      nrow = 2, byrow = TRUE
    ),
    sigma_nu    = c(nu_sd,    nu_sd / 2),
    sigma_alpha = c(alpha_sd, alpha_sd / 2),
    sigma_tau   = c(tau_sd,   tau_sd / 2)
  )
}

# simulation and fit
run_one <- function(I, J_per, true, model,
                    rhat_threshold     = 1.05,
                    hopeless_threshold = 10,
                    iter1 = 1000, warmup1 = 300,
                    iter2 = 3000, warmup2 = 1000) {

  x          <- rnorm(I)
  cond_dummy <- c(0, 1)
  J_arr      <- array(J_per, dim = c(I, K))
  C_arr      <- array(NA,    dim = c(I, K))
  MRT_mat    <- array(NA,    dim = c(I, K))
  VRT_mat    <- array(NA,    dim = c(I, K))
  true_nu    <- array(NA,    dim = c(I, K))
  true_alpha <- array(NA,    dim = c(I, K))
  true_tau   <- array(NA,    dim = c(I, K))

  for (i in 1:I) for (k in 1:K) {
    nu_i    <- rnorm(1, true$mu_nu    + x[i]*true$b1_nu + cond_dummy[k]*true$b2_nu, true$sigma_nu)
    alpha_i <- rnorm(1, true$mu_alpha + cond_dummy[k]*true$b_alpha,                  true$sigma_alpha)
    tau_i   <- rnorm(1, true$mu_tau   + cond_dummy[k]*true$b_tau,                    true$sigma_tau)

    nu_i    <- max(-3,   min(3,   nu_i))
    alpha_i <- max(0.1,  min(3.0, alpha_i))
    tau_i   <- max(0.05, tau_i)

    true_nu[i,k]    <- nu_i
    true_alpha[i,k] <- alpha_i
    true_tau[i,k]   <- tau_i

    u      <- exp(-alpha_i * nu_i)
    pi_c   <- 1 / (1 + u)
    mu_rt  <- tau_i + (alpha_i / (2*nu_i)) * ((1 - u) / (1 + u))
    sig2rt <- (alpha_i / (2*nu_i^3)) * ((1 - 2*alpha_i*nu_i*u - u^2) / (1+u)^2)

    C_i          <- max(rbinom(1, J_per, pi_c), 2)
    C_arr[i,k]   <- C_i
    MRT_mat[i,k] <- rnorm(1, mu_rt,  sqrt(sig2rt / C_i))
    VRT_mat[i,k] <- max(rnorm(1, sig2rt, sqrt(2*sig2rt^2 / (C_i-1))), 1e-6)
    J_arr[i,k]   <- J_per
  }

  ez     <- ez_point_estimates(C_arr, J_arr, MRT_mat, VRT_mat, I, K)
  priors <- derive_priors(ez)

  X_nu    <- array(NA, dim = c(I, K, 3))
  X_alpha <- array(NA, dim = c(I, K, 2))
  X_tau   <- array(NA, dim = c(I, K, 2))
  for (i in 1:I) for (k in 1:K) {
    X_nu[i,k,]    <- c(1, x[i], cond_dummy[k])
    X_alpha[i,k,] <- c(1, cond_dummy[k])
    X_tau[i,k,]   <- c(1, cond_dummy[k])
  }

  stan_data <- list(
    I = I, K = K, P_nu = 3, P_alpha = 2, P_tau = 2,
    J = J_arr, C = C_arr, MRT = MRT_mat, VRT = VRT_mat,
    X_nu = X_nu, X_alpha = X_alpha, X_tau = X_tau,
    prior_beta_nu     = priors$beta_nu,
    prior_beta_alpha  = priors$beta_alpha,
    prior_beta_tau    = priors$beta_tau,
    prior_sigma_nu    = priors$sigma_nu,
    prior_sigma_alpha = priors$sigma_alpha,
    prior_sigma_tau   = priors$sigma_tau
  )

  exclude_row <- function(reason) {
    data.frame(param=NA, true_val=NA, estimate=NA, q5=NA, q95=NA,
               condition=NA, level=NA, excluded=TRUE,
               reason=reason, I=I, J=J_per)
  }

  hopeless <- function(fit) {
    rhats <- tryCatch(summary(fit)$summary[, "Rhat"], error = function(e) NA)
    any(is.nan(rhats) | is.infinite(rhats) | rhats > hopeless_threshold,
        na.rm = FALSE)
  }

  converged <- function(fit) {
    rhats <- summary(fit)$summary[, "Rhat"]
    !any(rhats > rhat_threshold, na.rm = TRUE)
  }

  fit <- tryCatch(
    sampling(model, data = stan_data,
             chains=2, iter=iter1, warmup=warmup1,
             cores=1, refresh=0,
             control=list(adapt_delta=0.90)),
    error = function(e) NULL)

  if (is.null(fit)) {
    message("First fit crashed — skipping (I=", I, ", J=", J_per, ")")
    return(exclude_row("crash_1"))
  }

  if (hopeless(fit)) {
    max_rhat <- round(max(summary(fit)$summary[, "Rhat"], na.rm = TRUE), 3)
    message("Degenerate fit — excluding without refit (max Rhat=", max_rhat,
            ", I=", I, ", J=", J_per, ")")
    return(exclude_row("degenerate"))
  }

  if (!converged(fit)) {
    max_rhat <- round(max(summary(fit)$summary[, "Rhat"], na.rm=TRUE), 3)
    message("Refit triggered (max Rhat=", max_rhat,
            ", I=", I, ", J=", J_per, ")")

    fit <- tryCatch(
      sampling(model, data=stan_data,
               chains=4, iter=iter2, warmup=warmup2,
               cores=1, refresh=0,
               control=list(adapt_delta=0.95)),
      error = function(e) NULL)

    if (is.null(fit)) {
      message("Second fit crashed — excluding (I=", I, ", J=", J_per, ")")
      return(exclude_row("crash_2"))
    }

    if (!converged(fit)) {
      max_rhat <- round(max(summary(fit)$summary[, "Rhat"], na.rm=TRUE), 3)
      message("Still not converged after refit (max Rhat=", max_rhat,
              ") — excluding (I=", I, ", J=", J_per, ")")
      return(exclude_row("no_convergence"))
    }
  }

  summ <- summary(fit)$summary

  group_results <- data.frame(
    param     = c("mu_nu", "mu_alpha", "mu_tau"),
    true_val  = c(true$mu_nu, true$mu_alpha, true$mu_tau),
    estimate  = summ[c("beta_nu[1]", "beta_alpha[1]", "beta_tau[1]"), "mean"],
    q5        = summ[c("beta_nu[1]", "beta_alpha[1]", "beta_tau[1]"), "2.5%"],
    q95       = summ[c("beta_nu[1]", "beta_alpha[1]", "beta_tau[1]"), "97.5%"],
    condition = NA, level = "group", excluded = FALSE, reason = NA,
    I = I, J = J_per, row.names = NULL
  )

  indiv_results <- bind_rows(lapply(1:K, function(k) {
    nu_names    <- paste0("nu[",    1:I, ",", k, "]")
    alpha_names <- paste0("alpha[", 1:I, ",", k, "]")
    tau_names   <- paste0("tau[",   1:I, ",", k, "]")
    bind_rows(
      data.frame(param="nu",    true_val=true_nu[,k],
                 estimate=summ[nu_names,    "mean"],
                 q5=summ[nu_names,    "2.5%"], q95=summ[nu_names,    "97.5%"],
                 condition=k, level="individual", excluded=FALSE, reason=NA,
                 I=I, J=J_per),
      data.frame(param="alpha", true_val=true_alpha[,k],
                 estimate=summ[alpha_names, "mean"],
                 q5=summ[alpha_names, "2.5%"], q95=summ[alpha_names, "97.5%"],
                 condition=k, level="individual", excluded=FALSE, reason=NA,
                 I=I, J=J_per),
      data.frame(param="tau",   true_val=true_tau[,k],
                 estimate=summ[tau_names,   "mean"],
                 q5=summ[tau_names,   "2.5%"], q95=summ[tau_names,   "97.5%"],
                 condition=k, level="individual", excluded=FALSE, reason=NA,
                 I=I, J=J_per)
    )
  }))

  bind_rows(group_results, indiv_results)
}

# -------------------------------------------------------
# parallel simulation
# -------------------------------------------------------

jobs      <- expand.grid(I = I_vals, J = J_vals, sim = seq_len(N_sim))
jobs_list <- split(jobs, seq_len(nrow(jobs)))

n_workers <- detectCores() - 1
cat("Running", nrow(jobs), "jobs on", n_workers, "workers\n")

cl <- makeCluster(n_workers)

clusterSetRNGStream(cl, 42)

clusterExport(cl, varlist = c(
  "run_one", "ez_point_estimates", "derive_priors",
  "true", "K"
))

clusterEvalQ(cl, {
  library(rstan)
  library(dplyr)
  rstan_options(auto_write = TRUE)
  options(mc.cores = 1)
  ez_cond <<- stan_model("new_model_w_cond_matrices.stan")
})

results_list <- parLapply(cl, jobs_list, function(row) {
  run_one(
    I     = row$I,
    J_per = row$J,
    true  = true,
    model = ez_cond
  )
})

stopCluster(cl)

all_results <- bind_rows(results_list)
saveRDS(all_results, file = "simulation_results.rds")

# -------------------------------------------------------
# post-processing
# -------------------------------------------------------

excluded <- all_results |> filter(excluded == TRUE)
results  <- all_results |> filter(excluded == FALSE)

exclusion_table <- excluded |>
  group_by(I, J) |>
  summarise(n_excluded = n(), .groups = "drop") |>
  mutate(pct = round(n_excluded / N_sim * 100, 1)) |>
  pivot_wider(names_from=J, values_from=c(n_excluded, pct),
              names_glue="{.value}_J{J}", values_fill=0)

exclusion_report <- excluded |>
  group_by(I, J, reason) |>
  summarise(n = n(), pct = round(n() / N_sim * 100, 1), .groups="drop")

print(exclusion_report)

total_sims  <- N_sim * length(I_vals) * length(J_vals)
n_excluded  <- nrow(excluded)
pct_overall <- round(n_excluded / total_sims * 100, 1)

# -------------------------------------------------------
# prepare plotting
# -------------------------------------------------------

results <- results |>
  mutate(
    I_label   = factor(paste0("I = ", I), levels=paste0("I = ", I_vals)),
    J_label   = factor(paste0("J = ", J), levels=paste0("J = ", J_vals)),
    condition = ifelse(is.na(condition), NA, paste0("Condition ", condition))
  )

results_group <- results |> filter(level == "group")
results_indiv <- results |> filter(level == "individual")

dark_theme <- theme(
  plot.background  = element_rect(fill="black", colour=NA),
  plot.title       = element_text(colour="white", size=13, hjust=0.5),
  panel.background = element_rect(fill="black", colour=NA),
  panel.grid.major = element_line(colour="grey25", linewidth=0.3),
  panel.grid.minor = element_blank(),
  panel.border     = element_rect(colour="white", fill=NA, linewidth=0.4),
  strip.background = element_rect(fill="black", colour=NA),
  strip.text       = element_text(colour="white", size=10),
  axis.text        = element_text(colour="white", size=8),
  axis.ticks       = element_line(colour="white"),
  axis.title       = element_blank(),
  legend.position  = "none"
)

make_plot <- function(df, param_name, label, by_condition=FALSE) {
  p <- ggplot(df |> filter(param == param_name),
              aes(x=true_val, y=estimate)) +
    geom_abline(intercept=0, slope=1,
                linetype="dashed", colour="white", linewidth=0.5) +
    geom_smooth(method="lm", colour="goldenrod3",
                fill="goldenrod3", alpha=0.3, linewidth=0.7) +
    geom_point(colour="goldenrod", alpha=0.6, size=1.2) +
    labs(title=label) +
    dark_theme

  if (by_condition) {
    p + facet_grid(I_label + condition ~ J_label)
  } else {
    p + facet_grid(I_label ~ J_label)
  }
}

group_labels <- list(
  mu_nu    = expression(mu^{(nu)}),
  mu_alpha = expression(mu^{(alpha)}),
  mu_tau   = expression(mu^{(tau)})
)

plots_group <- lapply(names(group_labels), function(p)
  make_plot(results_group, p, group_labels[[p]], by_condition=FALSE))
names(plots_group) <- names(group_labels)

indiv_labels <- list(
  nu    = expression(nu),
  alpha = expression(alpha),
  tau   = expression(tau)
)

plots_indiv <- lapply(names(indiv_labels), function(p)
  make_plot(results_indiv, p, indiv_labels[[p]], by_condition=TRUE))
names(plots_indiv) <- names(indiv_labels)

plots_group[["mu_nu"]];  plots_group[["mu_alpha"]];  plots_group[["mu_tau"]]
plots_indiv[["nu"]];     plots_indiv[["alpha"]];     plots_indiv[["tau"]]

for (p in names(plots_group)) {
  ggsave(paste0("recovery_group_", p, ".png"),
         plot=plots_group[[p]], width=8, height=7, bg="black", dpi=300)
}
for (p in names(plots_indiv)) {
  ggsave(paste0("recovery_indiv_", p, ".png"),
         plot=plots_indiv[[p]], width=8, height=10, bg="black", dpi=300)
}
```




```{r check that stan uses the right arrays and not always the first one}

library(rstan)
library(ggplot2)
library(dplyr)
library(tidyr)

# small dataset 
I_test <- 15
J_test <- 100
K      <- 2

set.seed(99)

true <- list(
  mu_nu    = 1.0, b1_nu = 0.3, b2_nu = 0.5,
  mu_alpha = 1.5, b_alpha = 0.2,
  mu_tau   = 0.3, b_tau  = 0.02,
  sigma_nu = 0.3, sigma_alpha = 0.25, sigma_tau = 0.15
)

# generate data
x          <- rnorm(I_test)
cond_dummy <- c(0, 1)
J_arr      <- array(J_test, dim = c(I_test, K))
C_arr      <- array(NA, dim = c(I_test, K))
MRT_mat    <- array(NA, dim = c(I_test, K))
VRT_mat    <- array(NA, dim = c(I_test, K))
true_nu    <- array(NA, dim = c(I_test, K))
true_alpha <- array(NA, dim = c(I_test, K))
true_tau   <- array(NA, dim = c(I_test, K))

for (i in 1:I_test) for (k in 1:K) {
  nu_i    <- rnorm(1, true$mu_nu    + x[i]*true$b1_nu + cond_dummy[k]*true$b2_nu, true$sigma_nu)
  alpha_i <- rnorm(1, true$mu_alpha + cond_dummy[k]*true$b_alpha,                  true$sigma_alpha)
  tau_i   <- rnorm(1, true$mu_tau   + cond_dummy[k]*true$b_tau,                    true$sigma_tau)

  nu_i    <- max(-3,   min(3,   nu_i))
  alpha_i <- max(0.1,  min(3.0, alpha_i))
  tau_i   <- max(0.05, tau_i)

  true_nu[i,k]    <- nu_i
  true_alpha[i,k] <- alpha_i
  true_tau[i,k]   <- tau_i

  u      <- exp(-alpha_i * nu_i)
  pi_c   <- 1 / (1 + u)
  mu_rt  <- tau_i + (alpha_i / (2*nu_i)) * ((1 - u) / (1 + u))
  sig2rt <- (alpha_i / (2*nu_i^3)) * ((1 - 2*alpha_i*nu_i*u - u^2) / (1+u)^2)

  C_i            <- max(rbinom(1, J_test, pi_c), 2)
  C_arr[i,k]     <- C_i
  MRT_mat[i,k]   <- rnorm(1, mu_rt,  sqrt(sig2rt / C_i))
  VRT_mat[i,k]   <- max(rnorm(1, sig2rt, sqrt(2*sig2rt^2 / (C_i-1))), 1e-6)
}

X_nu    <- array(NA, dim = c(I_test, K, 3))
X_alpha <- array(NA, dim = c(I_test, K, 2))
X_tau   <- array(NA, dim = c(I_test, K, 2))
for (i in 1:I_test) for (k in 1:K) {
  X_nu[i,k,]    <- c(1, x[i], cond_dummy[k])
  X_alpha[i,k,] <- c(1, cond_dummy[k])
  X_tau[i,k,]   <- c(1, cond_dummy[k])
}

# derive priors from data
ez    <- ez_point_estimates(C_arr, J_arr, MRT_mat, VRT_mat, I_test, K)
priors <- derive_priors(ez)

stan_data <- list(
  I = I_test, K = K, P_nu = 3, P_alpha = 2, P_tau = 2,
  J = J_arr, C = C_arr, MRT = MRT_mat, VRT = VRT_mat,
  X_nu = X_nu, X_alpha = X_alpha, X_tau = X_tau,
  prior_beta_nu     = priors$beta_nu,
  prior_beta_alpha  = priors$beta_alpha,
  prior_beta_tau    = priors$beta_tau,
  prior_sigma_nu    = priors$sigma_nu,
  prior_sigma_alpha = priors$sigma_alpha,
  prior_sigma_tau   = priors$sigma_tau
)

# fit 
ez_cond <- stan_model("new_model_w_cond_matrices.stan")

fit_test <- sampling(ez_cond, data = stan_data,
                     chains = 2, iter = 2000, warmup = 500,
                     cores = 2, refresh = 0,
                     control = list(adapt_delta = 0.90))

summ <- summary(fit_test)$summary


# do all [i, k] cells have distinct estimates?
# extract posterior means for all cells

diag_df <- bind_rows(lapply(1:K, function(k) {
  bind_rows(lapply(1:I_test, function(i) {
    data.frame(
      i         = i,
      k         = k,
      condition = paste0("Condition ", k),
      true_nu    = true_nu[i, k],
      true_alpha = true_alpha[i, k],
      true_tau   = true_tau[i, k],
      est_nu     = summ[paste0("nu[",    i, ",", k, "]"), "mean"],
      est_alpha  = summ[paste0("alpha[", i, ",", k, "]"), "mean"],
      est_tau    = summ[paste0("tau[",   i, ",", k, "]"), "mean"]
    )
  }))
}))

# are all estimates distinct and not all collapsed to the same value?
cat("Check 1: all estimates distinct? \n")
cat("nu    unique estimates:", n_distinct(round(diag_df$est_nu,    4)), "/ expected", I_test * K, "\n")
cat("alpha unique estimates:", n_distinct(round(diag_df$est_alpha, 4)), "/ expected", I_test * K, "\n")
cat("tau   unique estimates:", n_distinct(round(diag_df$est_tau,   4)), "/ expected", I_test * K, "\n\n")


# condition 2 estimates systematically higher than condition 1?
cat("Check 2: condition shift in estimates \n")
cat("True b2_nu (condition shift on nu):", true$b2_nu, "\n")
cat("Observed shift (mean Cond2 - mean Cond1):\n")
diag_df |>
  group_by(condition) |>
  summarise(mean_nu    = mean(est_nu),
            mean_alpha = mean(est_alpha),
            mean_tau   = mean(est_tau),
            .groups = "drop") |>
  print()


# correlation between true and estimated values?

cat("Check 3: correlation true vs estimated \n")
cat("nu    correlation:", round(cor(diag_df$true_nu,    diag_df$est_nu),    3), "\n")
cat("alpha correlation:", round(cor(diag_df$true_alpha, diag_df$est_alpha), 3), "\n")
cat("tau   correlation:", round(cor(diag_df$true_tau,   diag_df$est_tau),   3), "\n\n")


# plot true vs estimated per condition

dark_theme <- theme(
  plot.background  = element_rect(fill = "black", colour = NA),
  plot.title       = element_text(colour = "white", size = 11, hjust = 0.5),
  panel.background = element_rect(fill = "black", colour = NA),
  panel.grid.major = element_line(colour = "grey25", linewidth = 0.3),
  panel.grid.minor = element_blank(),
  panel.border     = element_rect(colour = "white", fill = NA, linewidth = 0.4),
  strip.background = element_rect(fill = "black", colour = NA),
  strip.text       = element_text(colour = "white", size = 10),
  axis.text        = element_text(colour = "white", size = 8),
  axis.ticks       = element_line(colour = "white"),
  axis.title       = element_text(colour = "white", size = 9),
  legend.position  = "none"
)

make_check_plot <- function(true_col, est_col, title) {
  ggplot(diag_df, aes(x = .data[[true_col]], y = .data[[est_col]],
                      colour = condition)) +
    geom_abline(intercept = 0, slope = 1,
                linetype = "dashed", colour = "white", linewidth = 0.5) +
    geom_point(size = 2.5, alpha = 0.85) +
    geom_text(aes(label = i), colour = "white", size = 2.5,
              vjust = -0.8, hjust = 0.5) +
    scale_colour_manual(values = c("Condition 1" = "goldenrod",
                                   "Condition 2" = "steelblue1")) +
    facet_wrap(~ condition) +
    labs(title = title,
         x = "True value", y = "Posterior mean") +
    dark_theme +
    theme(legend.position = "none")
}

p_nu    <- make_check_plot("true_nu",    "est_nu",    expression(nu))
p_alpha <- make_check_plot("true_alpha", "est_alpha", expression(alpha))
p_tau   <- make_check_plot("true_tau",   "est_tau",   expression(tau))

p_nu
p_alpha
p_tau
```

```{r find out which priors were used}
set.seed(42)

prior_summary <- bind_rows(lapply(I_vals, function(I) {
  bind_rows(lapply(J_vals, function(J) {

    # generate one dataset
    x          <- rnorm(I)
    cond_dummy <- c(0, 1)
    J_arr      <- array(J, dim = c(I, K))
    C_arr      <- array(NA, dim = c(I, K))
    MRT_mat    <- array(NA, dim = c(I, K))
    VRT_mat    <- array(NA, dim = c(I, K))

    for (i in 1:I) for (k in 1:K) {
      nu_i    <- max(-3, min(3,   rnorm(1, true$mu_nu    + x[i]*true$b1_nu + cond_dummy[k]*true$b2_nu, true$sigma_nu)))
      alpha_i <- max(0.1, min(3.0, rnorm(1, true$mu_alpha + cond_dummy[k]*true$b_alpha, true$sigma_alpha)))
      tau_i   <- max(0.05, rnorm(1, true$mu_tau + cond_dummy[k]*true$b_tau, true$sigma_tau))

      u      <- exp(-alpha_i * nu_i)
      pi_c   <- 1 / (1 + u)
      mu_rt  <- tau_i + (alpha_i / (2*nu_i)) * ((1 - u) / (1 + u))
      sig2rt <- (alpha_i / (2*nu_i^3)) * ((1 - 2*alpha_i*nu_i*u - u^2) / (1+u)^2)

      C_i          <- max(rbinom(1, J, pi_c), 2)
      C_arr[i,k]   <- C_i
      MRT_mat[i,k] <- rnorm(1, mu_rt,  sqrt(sig2rt / C_i))
      VRT_mat[i,k] <- max(rnorm(1, sig2rt, sqrt(2*sig2rt^2 / (C_i-1))), 1e-6)
    }

    ez     <- ez_point_estimates(C_arr, J_arr, MRT_mat, VRT_mat, I, K)
    priors <- derive_priors(ez)

    data.frame(
      I = I, J = J,
      # intercept priors (means)
      prior_mean_nu    = round(priors$beta_nu[1, 1],    3),
      prior_mean_alpha = round(priors$beta_alpha[1, 1], 3),
      prior_mean_tau   = round(priors$beta_tau[1, 1],   3),
      # intercept priors (SDs)
      prior_sd_nu      = round(priors$beta_nu[2, 1],    3),
      prior_sd_alpha   = round(priors$beta_alpha[2, 1], 3),
      prior_sd_tau     = round(priors$beta_tau[2, 1],   3),
      # sigma priors
      prior_sigma_nu_mean    = round(priors$sigma_nu[1],    3),
      prior_sigma_alpha_mean = round(priors$sigma_alpha[1], 3),
      prior_sigma_tau_mean   = round(priors$sigma_tau[1],   3)
    )
  }))
}))

print(prior_summary)

# table 
kable(prior_summary,
      caption = "Data-derived prior hyperparameters for a representative dataset
                 per I x J condition. Prior means for intercepts are the grand
                 mean of EZ point estimates; SDs are twice the SD of EZ estimates.",
      col.names = c("I", "J",
                    "mu_nu mean", "mu_alpha mean", "mu_tau mean",
                    "mu_nu SD",   "mu_alpha SD",   "mu_tau SD",
                    "sigma_nu",   "sigma_alpha",   "sigma_tau"),
      align = "c")
```

full code am ende sollte haben
- big simulation including data derived priors and random slope
- violin plots for means and recovery plots mit reference line for individual parameters
- Data-derived priors ausgeben damit man sehen kann welche benutzt wurden 
- other measures for good convergence?



```{r big simulation with random slope and hopefully everything fixed}
library(rstan)
library(bayesplot)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(parallel)

# -------------------------------------------------------
# design 
# -------------------------------------------------------
I_vals <- c(20, 60, 150)
J_vals <- c(20, 60, 150)
N_sim  <- 5 #50
K      <- 2

# true parameter values
true <- list(
  mu_nu    = 1.0, b1_nu = 0.3, b2_nu = 0.5,
  mu_alpha = 1.5, b_alpha = 0.2,
  mu_tau   = 0.3, b_tau  = 0.02,
  sigma_nu = 0.3, sigma_alpha = 0.25, sigma_tau = 0.15,
  sigma_u  = 0.15
)

# -------------------------------------------------------
# EZ inverse equations
# -------------------------------------------------------
ez_point_estimates <- function(C_arr, J_arr, MRT_mat, VRT_mat, I, K) {
  nu_hat    <- matrix(NA, I, K)
  alpha_hat <- matrix(NA, I, K)
  tau_hat   <- matrix(NA, I, K)

  for (i in 1:I) for (k in 1:K) {
    J_ik  <- J_arr[i, k]
    Pc    <- C_arr[i, k] / J_ik
    Pc    <- max(min(Pc, 1 - 1 / (2 * J_ik)), 1 / (2 * J_ik))
    VRT   <- VRT_mat[i, k]
    MRT   <- MRT_mat[i, k]

    L      <- log(Pc / (1 - Pc))
    inside <- L * (Pc^2 * L - Pc * L + Pc - 0.5) / VRT

    if (is.na(inside) || inside <= 0) next

    nu_ik    <- sign(Pc - 0.5) * inside^(1/4)
    if (abs(nu_ik) < 1e-6) next

    alpha_ik <- L / nu_ik
    ev       <- exp(-alpha_ik * nu_ik)
    tau_ik   <- MRT - (alpha_ik / (2 * nu_ik)) * ((1 - ev) / (1 + ev))

    nu_hat[i, k]    <- nu_ik
    alpha_hat[i, k] <- alpha_ik
    tau_hat[i, k]   <- tau_ik
  }

  list(nu = nu_hat, alpha = alpha_hat, tau = tau_hat)
}

# -------------------------------------------------------
# derive priors from EZ point estimates
# -------------------------------------------------------
derive_priors <- function(ez) {
  nu_vals    <- as.vector(ez$nu)
  alpha_vals <- as.vector(ez$alpha)
  tau_vals   <- as.vector(ez$tau)

  nu_vals    <- nu_vals[!is.na(nu_vals) & nu_vals > -3   & nu_vals < 3]
  alpha_vals <- alpha_vals[!is.na(alpha_vals) & alpha_vals > 0.1 & alpha_vals < 3]
  tau_vals   <- tau_vals[!is.na(tau_vals) & tau_vals > 0.05]

  nu_sd    <- sd(nu_vals)
  alpha_sd <- sd(alpha_vals)
  tau_sd   <- sd(tau_vals)

  list(
    beta_nu = matrix(
      c(mean(nu_vals), 0,         0,
        nu_sd * 2,     nu_sd * 2, nu_sd * 2),
      nrow = 2, byrow = TRUE
    ),
    beta_alpha = matrix(
      c(mean(alpha_vals), 0,
        alpha_sd * 2,     alpha_sd * 2),
      nrow = 2, byrow = TRUE
    ),
    beta_tau = matrix(
      c(mean(tau_vals), 0,
        tau_sd * 2,     tau_sd * 2),
      nrow = 2, byrow = TRUE
    ),
    sigma_nu    = c(nu_sd,        nu_sd / 2),
    sigma_alpha = c(alpha_sd,     alpha_sd / 2),
    sigma_tau   = c(tau_sd,       tau_sd / 2),
    sigma_u     = c(nu_sd * 0.3,  nu_sd)
  )
}

# -------------------------------------------------------
# simulation and fit
# -------------------------------------------------------
run_one <- function(I, J_per, true, model,
                    rhat_threshold     = 1.05,
                    hopeless_threshold = 10,
                    iter1 = 1000, warmup1 = 300,
                    iter2 = 3000, warmup2 = 1000) {

  x          <- rnorm(I)
  cond_dummy <- c(0, 1)
  J_arr      <- array(J_per, dim = c(I, K))
  C_arr      <- array(NA,    dim = c(I, K))
  MRT_mat    <- array(NA,    dim = c(I, K))
  VRT_mat    <- array(NA,    dim = c(I, K))
  true_nu    <- array(NA,    dim = c(I, K))
  true_alpha <- array(NA,    dim = c(I, K))
  true_tau   <- array(NA,    dim = c(I, K))

  true_u <- rnorm(I, 0, true$sigma_u)

  for (i in 1:I) for (k in 1:K) {
    nu_i    <- rnorm(1, true$mu_nu + (true$b1_nu + true_u[i]) * x[i] +
                          cond_dummy[k] * true$b2_nu, true$sigma_nu)
    alpha_i <- rnorm(1, true$mu_alpha + cond_dummy[k]*true$b_alpha, true$sigma_alpha)
    tau_i   <- rnorm(1, true$mu_tau   + cond_dummy[k]*true$b_tau,   true$sigma_tau)

    nu_i    <- max(-3,   min(3,   nu_i))
    alpha_i <- max(0.1,  min(3.0, alpha_i))
    tau_i   <- max(0.05, tau_i)

    true_nu[i,k]    <- nu_i
    true_alpha[i,k] <- alpha_i
    true_tau[i,k]   <- tau_i

    ev     <- exp(-alpha_i * nu_i)
    pi_c   <- 1 / (1 + ev)
    mu_rt  <- tau_i + (alpha_i / (2*nu_i)) * ((1 - ev) / (1 + ev))
    sig2rt <- (alpha_i / (2*nu_i^3)) * ((1 - 2*alpha_i*nu_i*ev - ev^2) / (1+ev)^2)

    C_i          <- max(rbinom(1, J_per, pi_c), 2)
    C_arr[i,k]   <- C_i
    MRT_mat[i,k] <- rnorm(1, mu_rt,  sqrt(sig2rt / C_i))
    VRT_mat[i,k] <- max(rnorm(1, sig2rt, sqrt(2*sig2rt^2 / (C_i-1))), 1e-6)
    J_arr[i,k]   <- J_per
  }

  ez     <- ez_point_estimates(C_arr, J_arr, MRT_mat, VRT_mat, I, K)
  priors <- derive_priors(ez)

  X_nu    <- array(NA, dim = c(I, K, 3))
  X_alpha <- array(NA, dim = c(I, K, 2))
  X_tau   <- array(NA, dim = c(I, K, 2))
  for (i in 1:I) for (k in 1:K) {
    X_nu[i,k,]    <- c(1, x[i], cond_dummy[k])
    X_alpha[i,k,] <- c(1, cond_dummy[k])
    X_tau[i,k,]   <- c(1, cond_dummy[k])
  }

  stan_data <- list(
    I = I, K = K, P_nu = 3, P_alpha = 2, P_tau = 2,
    J = J_arr, C = C_arr, MRT = MRT_mat, VRT = VRT_mat,
    X_nu = X_nu, X_alpha = X_alpha, X_tau = X_tau,
    x = x,
    prior_beta_nu     = priors$beta_nu,
    prior_beta_alpha  = priors$beta_alpha,
    prior_beta_tau    = priors$beta_tau,
    prior_sigma_nu    = priors$sigma_nu,
    prior_sigma_alpha = priors$sigma_alpha,
    prior_sigma_tau   = priors$sigma_tau,
    prior_sigma_u     = priors$sigma_u
  )

  exclude_row <- function(reason) {
    data.frame(param=NA, true_val=NA, estimate=NA, q5=NA, q95=NA,
               condition=NA, level=NA, excluded=TRUE,
               reason=reason, I=I, J=J_per,
               prior_mean=NA, prior_sd=NA)
  }

  hopeless <- function(fit) {
    rhats <- tryCatch(summary(fit)$summary[, "Rhat"], error = function(e) NA)
    any(is.nan(rhats) | is.infinite(rhats) | rhats > hopeless_threshold,
        na.rm = FALSE)
  }

  converged <- function(fit) {
    rhats <- summary(fit)$summary[, "Rhat"]
    !any(rhats > rhat_threshold, na.rm = TRUE)
  }

  fit <- tryCatch(
    sampling(model, data = stan_data,
             chains=2, iter=iter1, warmup=warmup1,
             cores=1, refresh=0,
             control=list(adapt_delta=0.90)),
    error = function(e) NULL)

  if (is.null(fit)) {
    message("First fit crashed — skipping (I=", I, ", J=", J_per, ")")
    return(exclude_row("crash_1"))
  }

  if (hopeless(fit)) {
    max_rhat <- round(max(summary(fit)$summary[, "Rhat"], na.rm = TRUE), 3)
    message("Degenerate fit — excluding without refit (max Rhat=", max_rhat,
            ", I=", I, ", J=", J_per, ")")
    return(exclude_row("degenerate"))
  }

  if (!converged(fit)) {
    max_rhat <- round(max(summary(fit)$summary[, "Rhat"], na.rm=TRUE), 3)
    message("Refit triggered (max Rhat=", max_rhat,
            ", I=", I, ", J=", J_per, ")")

    fit <- tryCatch(
      sampling(model, data=stan_data,
               chains=4, iter=iter2, warmup=warmup2,
               cores=1, refresh=0,
               control=list(adapt_delta=0.95)),
      error = function(e) NULL)

    if (is.null(fit)) {
      message("Second fit crashed — excluding (I=", I, ", J=", J_per, ")")
      return(exclude_row("crash_2"))
    }

    if (!converged(fit)) {
      max_rhat <- round(max(summary(fit)$summary[, "Rhat"], na.rm=TRUE), 3)
      message("Still not converged after refit (max Rhat=", max_rhat,
              ") — excluding (I=", I, ", J=", J_per, ")")
      return(exclude_row("no_convergence"))
    }
  }

  summ <- summary(fit)$summary

  # save derived intercept priors alongside estimates for reporting
  group_results <- data.frame(
    param     = c("mu_nu", "mu_alpha", "mu_tau"),
    true_val  = c(true$mu_nu, true$mu_alpha, true$mu_tau),
    estimate  = summ[c("beta_nu[1]", "beta_alpha[1]", "beta_tau[1]"), "mean"],
    q5        = summ[c("beta_nu[1]", "beta_alpha[1]", "beta_tau[1]"), "2.5%"],
    q95       = summ[c("beta_nu[1]", "beta_alpha[1]", "beta_tau[1]"), "97.5%"],
    condition = NA, level = "group", excluded = FALSE, reason = NA,
    I = I, J = J_per,
    prior_mean = c(priors$beta_nu[1,1], priors$beta_alpha[1,1], priors$beta_tau[1,1]),
    prior_sd   = c(priors$beta_nu[2,1], priors$beta_alpha[2,1], priors$beta_tau[2,1]),
    row.names  = NULL
  )

  indiv_results <- bind_rows(lapply(1:K, function(k) {
    nu_names    <- paste0("nu[",    1:I, ",", k, "]")
    alpha_names <- paste0("alpha[", 1:I, ",", k, "]")
    tau_names   <- paste0("tau[",   1:I, ",", k, "]")
    bind_rows(
      data.frame(param="nu",    true_val=true_nu[,k],
                 estimate=summ[nu_names,    "mean"],
                 q5=summ[nu_names,    "2.5%"], q95=summ[nu_names,    "97.5%"],
                 condition=k, level="individual", excluded=FALSE, reason=NA,
                 I=I, J=J_per, prior_mean=NA, prior_sd=NA),
      data.frame(param="alpha", true_val=true_alpha[,k],
                 estimate=summ[alpha_names, "mean"],
                 q5=summ[alpha_names, "2.5%"], q95=summ[alpha_names, "97.5%"],
                 condition=k, level="individual", excluded=FALSE, reason=NA,
                 I=I, J=J_per, prior_mean=NA, prior_sd=NA),
      data.frame(param="tau",   true_val=true_tau[,k],
                 estimate=summ[tau_names,   "mean"],
                 q5=summ[tau_names,   "2.5%"], q95=summ[tau_names,   "97.5%"],
                 condition=k, level="individual", excluded=FALSE, reason=NA,
                 I=I, J=J_per, prior_mean=NA, prior_sd=NA)
    )
  }))

  u_names <- paste0("u[", 1:I, "]")
  u_results <- data.frame(
    param     = "u",
    true_val  = true_u,
    estimate  = summ[u_names, "mean"],
    q5        = summ[u_names, "2.5%"],
    q95       = summ[u_names, "97.5%"],
    condition = NA, level = "random_slope",
    excluded  = FALSE, reason = NA,
    I = I, J = J_per, prior_mean = NA, prior_sd = NA,
    row.names = NULL
  )

  bind_rows(group_results, indiv_results, u_results)
}

# -------------------------------------------------------
# parallel simulation
# -------------------------------------------------------

jobs      <- expand.grid(I = I_vals, J = J_vals, sim = seq_len(N_sim))
jobs_list <- split(jobs, seq_len(nrow(jobs)))

cache_file <- paste0(tools::file_path_sans_ext(
  "new_model_w_cond_and_random_slope.stan"), ".rds")
if (file.exists(cache_file)) file.remove(cache_file)

n_workers <- detectCores() - 1
cat("Running", nrow(jobs), "jobs on", n_workers, "workers\n")

cl <- makeCluster(n_workers)
clusterSetRNGStream(cl, 42)

clusterExport(cl, varlist = c(
  "run_one", "ez_point_estimates", "derive_priors",
  "true", "K"
))

clusterEvalQ(cl, {
  library(rstan)
  library(dplyr)
  rstan_options(auto_write = TRUE)
  options(mc.cores = 1)
  ez_cond <<- stan_model("new_model_w_cond_and_random_slope.stan")
})

results_list <- parLapply(cl, jobs_list, function(row) {
  run_one(
    I     = row$I,
    J_per = row$J,
    true  = true,
    model = ez_cond
  )
})

stopCluster(cl)

all_results <- bind_rows(results_list)
saveRDS(all_results, file = "simulation_results.rds")

# -------------------------------------------------------
# post-processing
# -------------------------------------------------------

excluded <- all_results |> filter(excluded == TRUE)
results  <- all_results |> filter(excluded == FALSE)

exclusion_report <- excluded |>
  group_by(I, J, reason) |>
  summarise(n = n(), pct = round(n() / N_sim * 100, 1), .groups="drop")
print(exclusion_report)

total_sims  <- N_sim * length(I_vals) * length(J_vals)
n_excluded  <- nrow(excluded)
pct_overall <- round(n_excluded / total_sims * 100, 1)

results <- results |>
  mutate(
    I_label   = factor(paste0("I = ", I), levels=paste0("I = ", I_vals)),
    J_label   = factor(paste0("J = ", J), levels=paste0("J = ", J_vals)),
    condition = ifelse(is.na(condition), NA, paste0("Condition ", condition))
  )

results_group  <- results |> filter(level == "group")
results_indiv  <- results |> filter(level == "individual")
results_u      <- results |> filter(level == "random_slope")

# -------------------------------------------------------
# data-derived prior summary table
# -------------------------------------------------------

prior_summary <- results_group |>
  group_by(param, I, J) |>
  summarise(
    prior_mean_avg = round(mean(prior_mean, na.rm=TRUE), 3),
    prior_mean_sd  = round(sd(prior_mean,   na.rm=TRUE), 3),
    prior_sd_avg   = round(mean(prior_sd,   na.rm=TRUE), 3),
    prior_sd_sd    = round(sd(prior_sd,     na.rm=TRUE), 3),
    .groups = "drop"
  ) |>
  mutate(
    prior_mean_fmt = paste0(prior_mean_avg, " (", prior_mean_sd, ")"),
    prior_sd_fmt   = paste0(prior_sd_avg,   " (", prior_sd_sd,   ")")
  )

cat("\n--- Data-derived priors (mean across simulations, SD in brackets) ---\n")
print(prior_summary |> select(param, I, J, prior_mean_fmt, prior_sd_fmt))

kable(
  prior_summary |>
    select(param, I, J, prior_mean_fmt, prior_sd_fmt) |>
    rename(Parameter = param,
           `Prior mean` = prior_mean_fmt,
           `Prior SD`   = prior_sd_fmt),
  caption = "Data-derived prior hyperparameters for the intercept of each
             group-level parameter, averaged across simulations per I x J
             condition. Values in brackets are the SD across simulations,
             reflecting how much the derived priors vary with the data."
)

# -------------------------------------------------------
# shared plot theme
# -------------------------------------------------------

dark_theme <- theme(
  plot.background  = element_rect(fill="black", colour=NA),
  plot.title       = element_text(colour="white", size=13, hjust=0.5),
  panel.background = element_rect(fill="black", colour=NA),
  panel.grid.major = element_line(colour="grey25", linewidth=0.3),
  panel.grid.minor = element_blank(),
  panel.border     = element_rect(colour="white", fill=NA, linewidth=0.4),
  strip.background = element_rect(fill="black", colour=NA),
  strip.text       = element_text(colour="white", size=10),
  axis.text        = element_text(colour="white", size=8),
  axis.ticks       = element_line(colour="white"),
  axis.title       = element_text(colour="white", size=9),
  legend.position  = "none"
)

# -------------------------------------------------------
# violin plots for group-level means (bias distributions)
# -------------------------------------------------------

plot_group_bias <- function(param_name, label) {
  df <- results_group |>
    filter(param == param_name) |>
    mutate(bias = estimate - true_val)

  ggplot(df, aes(x = J_label, y = bias)) +
    geom_hline(yintercept = 0,
               linetype = "dashed", colour = "white", linewidth = 0.5) +
    geom_violin(fill = "goldenrod", colour = NA,
                alpha = 0.35, trim = TRUE, scale = "width") +
    geom_jitter(colour = "goldenrod", alpha = 0.7,
                size = 1.4, width = 0.08) +
    stat_summary(fun = mean, geom = "crossbar",
                 colour = "white", linewidth = 0.6,
                 width = 0.4, fatten = 1) +
    facet_grid(I_label ~ .) +
    labs(title = label,
         x = NULL,
         y = "Bias (estimate \u2212 true)") +
    dark_theme
}

bias_labels <- list(
  mu_nu    = expression(mu^{(nu)}),
  mu_alpha = expression(mu^{(alpha)}),
  mu_tau   = expression(mu^{(tau)})
)

plots_bias <- lapply(names(bias_labels), function(p)
  plot_group_bias(p, bias_labels[[p]]))
names(plots_bias) <- names(bias_labels)

# -------------------------------------------------------
# recovery scatter plots for individual parameters
# -------------------------------------------------------

make_recovery_plot <- function(df, param_name, label, by_condition=FALSE) {
  p <- ggplot(df |> filter(param == param_name),
              aes(x = true_val, y = estimate)) +
    geom_abline(intercept = 0, slope = 1,
                linetype = "dashed", colour = "white", linewidth = 0.5) +
    geom_smooth(method = "lm", colour = "goldenrod3",
                fill = "goldenrod3", alpha = 0.3, linewidth = 0.7) +
    geom_point(colour = "goldenrod", alpha = 0.6, size = 1.2) +
    labs(title = label) +
    dark_theme +
    theme(axis.title = element_blank())

  if (by_condition) {
    p + facet_grid(I_label + condition ~ J_label)
  } else {
    p + facet_grid(I_label ~ J_label)
  }
}

indiv_labels <- list(
  nu    = expression(nu),
  alpha = expression(alpha),
  tau   = expression(tau)
)

plots_indiv <- lapply(names(indiv_labels), function(p)
  make_recovery_plot(results_indiv, p, indiv_labels[[p]], by_condition=TRUE))
names(plots_indiv) <- names(indiv_labels)

plot_u <- make_recovery_plot(results_u, "u", expression(u[i]),
                             by_condition=FALSE)

# -------------------------------------------------------
# view all plots
# -------------------------------------------------------

plots_bias[["mu_nu"]]
plots_bias[["mu_alpha"]]
plots_bias[["mu_tau"]]

plots_indiv[["nu"]]
plots_indiv[["alpha"]]
plots_indiv[["tau"]]

plot_u

# -------------------------------------------------------
# save all plots
# -------------------------------------------------------

for (p in names(plots_bias)) {
  ggsave(paste0("bias_group_", p, ".png"),
         plot = plots_bias[[p]], width = 5, height = 7,
         bg = "black", dpi = 300)
}

for (p in names(plots_indiv)) {
  ggsave(paste0("recovery_indiv_", p, ".png"),
         plot = plots_indiv[[p]], width = 8, height = 10,
         bg = "black", dpi = 300)
}

ggsave("recovery_u.png", plot = plot_u,
       width = 8, height = 7, bg = "black", dpi = 300)
```



