---
title: "Demo Application in JAGS using the Krakow Dataset (Grabowska et al., 2025)"
author: "Adriana F. Chávez De la Peña"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: false
    lightbox: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 7)
library(R2jags)
library(here)
```

This document reproduces Anne's real-data analysis of the Krakow dataset (Grabowska et al., 2025) in JAGS. Anne's STAN implementation can be found in:

- `real data scripts and models/grabowska_ez_fit_krakow.R` (main fit script)
- `real data scripts and models/grabowska_ez_model.stan` (model)

The goal of this applied demo is to fit a hierarchical Bayesian EZ-DDM using design matrices to the data collected by Grabowska et al. (2025).

# About the data (Krakow dataset description) 

Here, we present a brief overview of the original study from which the data comes from (Grabowska et al., 2025)

## The experiment 

Participants completed a modified version of the **speeded arrow flanker task** as part of a large EEG project on post-response adaptation. The number of participants in the sample after cleaning was $N=222$ (participants who made too few errors or whose EEG data had poor quality were excluded from the analysis).

On each trial, participants' task was to **indicate whether a target arrow shown on screen was pointing left or right**. These target arrows appeared between to ``flanker'' arrows, which were either pointing in the same direction as the target (**congruent trials**, $70\%$ of trials) or in the opposite direction (**incongruent trials**, $30\%$ of trials). Importantly, on every trial participants only had a 600ms window to respond (hence why this is referred to as a ***speeded* arrow flanker task**). Participants complete a total of $T=300$ trials across 6 blocks.

## Study focus

This study focuses on **post-response adaptation**: Does the previous trial outcome (correct vs. error response) and the congruency condition (congruent vs. incongruent) have an effect on subsequent choices? 

# About this demo

We work with the data file `krakow_data_standardized.csv`, available for download [here](https://osf.io/36pvg/overview). The file is stored locally under `JAGS/demos/applications/data/`.

```{r}
dataFile <- here("JAGS", "demos", "applications", "data",
                 "krakow_data_standardized.csv")
```

We aggregate each participant's trials into **4 cells** following a $2 \times 2$ design defined by condition (congruent vs. incongruent) $\times$ past outcome (post-correct vs. post-error). Within each factor, levels are contrast-coded as $\pm 1$:

| Cell $k$ | Condition | Past outcome | Coding $(\text{Cond} \times \text{Past})$ |
|----------|-----------|---------------|--------------------------------------|
| 1 | Congruent | Post-correct | $(+1, +1)$ |
| 2 | Congruent | Post-error | $(+1, -1)$ |
| 3 | Incongruent | Post-correct | $(-1, +1)$ |
| 4 | Incongruent | Post-error | $(-1, -1)$ |

We do **not** include the EEG predictors from Grabowska et al. (same choice as Anne's main Stan fit).

## Design matrix specification

The model uses three **3D design arrays** to specify predictors for each DDM parameter. Each array has dimensions $I \times K \times P$, where for each participant $i$ and cell $k$, the slice $X[i,k,]$ is a row vector containing $P$ predictor values.

| Parameter | Design Array | Dimensions | Predictors |
|-----------|--------------|------------|------------|
| $\nu$ (drift) | $X_\nu[i,k,p]$ | $I \times K \times P_\nu$ | Intercept + Condition + Past + Condition $\times$ Past |
| $\alpha$ (boundary) | $X_\alpha[i,k,p]$ | $I \times K \times P_\alpha$ | Intercept + Past |
| $\tau$ (non-decision) | $X_\tau[i,k,p]$ | $I \times K \times P_\tau$ | Intercept only |

Where:

- $I$ = number of participants
- $K$ = number of within-subject cells
- $P$ = number of predictors (including intercept) for each parameter:
  - $P_\nu = 4$: Intercept + Condition + Past + Condition $\times$ Past
  - $P_\alpha = 2$: Intercept + Past
  - $P_\tau = 1$: Intercept only (no Condition or Past trial effects)

```{r}
K <- 4        
P_nu    <- 4 
P_alpha <- 2 
P_tau   <- 1 
```

Using the same $I \times K \times P$ structure for all design arrays $X_\tau$, $X_\alpha$, and $X_\nu$ ensures uniform indexing in the JAGS model that supports future extensions that include additional predictors. These fixed effects can be easily incorporated by using inner products (e.g., $\mu_{\tau,ik} = X_\tau[i,k,] \cdot \beta_\tau$). 

## The model 

In this demo, we use a model with **person-specific random slopes** on the columns of $X_\nu$ and $X_\alpha$ that correspond to the predictors of interest (congruency condition, past trial outcome, and their interaction). These random effects need to be added directly into the JAGS model as individual slopes.

```{r}
# Define design matrix for each cell
# Columns: [Intercept, Condition, Past, Cond x Past]
cell_design <- matrix(
    c( 1,  1,  1,  1,
       1,  1, -1, -1,
       1, -1,  1, -1,
       1, -1, -1,  1),
    nrow = K, byrow = TRUE
)
```

